\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2020-01-16 analysis/paper/paper.tex
fixedMaxComments, added questions, averaged absolute risks
I fixed a few of Max's comments, but some I was not certain about. particularly what to do with CFC's own package methods exploration. they dove into aspects I did not attempt to cover in the casebase lit review. NOTE: cmprsk was found in initial scraper, but was ignored in the manual pass through.

The final plot no longer uses a single individual but averages all individuals in our dataset. still uses our work around. Next I will attempt to use riskRegression and compare.

need to add timereg (riskregression is a wrapper for timereg)

At the time of writing, a cursory examination of CRAN's task view on
survival analysis reveals that there are over 250 packages related to
survival analysis \citeyearpar{survTaskView}. For the purposes of this
article, we restricted our description to packages that implement at
least one of the following features: parametric modeling,
non-proportional hazard models, competing risk analysis, penalized
estimation, and cumulative incidence curve estimation. By searching for
appropriate keywords in the \code{DESCRIPTION} file of these packages,
we found 60 relevant packages. These 60 packages were then manually
examined to determine which ones are comparable to \pkg{casebase}. In
particular, we excluded packages that were focused on a different set of
problems, such as frailty and multi-state models. The remaining 14
packages appear in Table \ref{tab:surv-pkgs}, along with a description
of some of the functionalities they offer.

Several packages implement penalized estimation for the Cox model:
\pkg{glmnet} \citeyearpar{regpathcox}, \pkg{glmpath}
\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
\pkg{RiskRegression} \citeyearpar{gerds_blanche}. Moreover, some
packages also include penalized estimation in the context of Cox models
with time-varying coefficients: \pkg{CoxRidge}
\citeyearpar{perperoglou}, \pkg{rstpm2} \citeyearpar{clements_liu} and
\pkg{survival} \citeyearpar{survival-package}. On the other hand,
\pkg{casebase} provides penalized estimation in the context of
parametric hazards. To our knowledge, this is the only package to offer
this functionality.
\textcolor{red}{Is this true? Because it seems to be a logical conclusion from your paragraph.}

Parametric survival models are implemented in a handful of packages:
\pkg{CFC} \citeyearpar{sharabiani_mahani_2019}, \pkg{flexsurv}
\citeyearpar{flexsurv}, \pkg{SmoothHazard} \citeyearpar{smoothHazard},
\pkg{rsptm2} \citeyearpar{clements_liu}, \pkg{mets}
\citeyearpar{scheike2014estimating}, and \pkg{survival}. The types of
models they allow vary for each package. For example, \pkg{SmoothHazard}
is limited to Weibull distributions \citeyearpar{smoothHazard}, whereas
both \pkg{flexsurv} and \pkg{survival} allow users to supply any
distribution of their choice. Also, \pkg{flexsurv}, \pkg{smoothhazard},
\pkg{mets} and \pkg{rstpm2} also have the ability to model the effect of
time using splines, which allows flexible modeling of the hazard
function. Moreover, \pkg{flexsurv} has the ability to estimate both
scale and shape parameters for a variety of parametric families (see
Table \ref{tab:dists}). As discussed above, \pkg{casebase} can model any
parametric family whose log-hazard can be expressed as a linear model of
covariates (including time). Therefore, our package allows the user to
model the effect of time using splines, and through interaction terms
involving covariates and time, it also allows user to fit time-varying
coefficient models. However, we do not explicitly model any shape
parameter, unlike \pkg{flexsurv}.

Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv},
\pkg{mets} and \pkg{survival} contain implementations for competing
risks. The differentiating factor between these packages and
\pkg{casebase} is that \pkg{casebase} handles competing risks through
multiple logistic regression.
\textcolor{red}{There's a literature review of packages for competing-risk analysis in the paper for \pkg{CFC}. Could you incorporate some of their discussion, with proper citations? The biggest omission right now is \pkg{cmprsk}, which we actually use below.}

@@

\begin{CodeInput}
R> library(casebase)
R> data(bmtcrr)
\end{CodeInput}
\end{CodeChunk}

@@

For the third case study, we will look at regularized estimation for the
hazard function. This can be performed using a combination of case-base
sampling and regularized logistic regression. To demonstrate this
capability, we examine the Study to Understand Prognoses Preferences
Outcomes and Risks of Treatment (SUPPORT) dataset
\citep{knaus1995support}. The SUPPORT dataset tracks death for
individuals who are considered seriously ill within a hospital.
Imputation was conducted using predictive mean matching and using the
\pkg{mice} package in \proglang{R} \citep{mice}. Before imputation,
there was a total of 1000 individuals and 35 variables. After
imputation, we have 474 individuals.
\textcolor{red}{How come we lose individuals during imputation?} Of
these, 67.5\% experienced the event of interest. A description of each
variable can be found in Table \ref{tab:support1} and a breakdown of
each categorical variable in Table \ref{tab:support2}. The original data
was obtained from the website of the Department of Biostatistics at
Vanderbilt University. The imputed dataset is available as part of the
\pkg{casebase} package.

@@

As before, the first step is to visualize the incidence density. The
population-time plot for the SUPPORT data appears in Figure
\ref{fig:support-poptime}.
\textcolor{red}{This population-time plot is strange. The red dots never seem to go above 200.}

@@

\begin{CodeInput}
R> data(support)
R> support_popTime <- popTime(support,
R+                            time = "d.time", event = "death")
R> plot(support_popTime)
\end{CodeInput}
\begin{figure}

@@

@@

\begin{CodeInput}
R> support <- dplyr::select(support, -hospdead)
R> 
R> # Randomly select one observation
R> indiv <- sample(1:nrow(support), size = 1)
\end{CodeInput}
\end{CodeChunk}

Our main objective is to estimate the cumulative incidence of death for
a given covariate profile. Given the number of covariates available, we
will do this using regularized estimation. As \pkg{casebase} makes use
of the \pkg{glmnet} package, we will use the function
\code{fitsmoothhazard.fit}, which follows the syntax from \code{glmnet}.
Specifically, we input a matrix \code{y}, which contains the time and
event variables, and a matrix \code{x}, which contains all other
variables we would like to include in the model. For the SUPPORT
dataset, all factor variables were converted into dummy variables.

@@

\begin{CodeInput}
R> # Create design matrix without the time variable
R> support_matrix <- model.matrix(death ~ . - d.time, 
R+                                data = support)
R> # Remove intercept term
R> support_matrix <- support_matrix[,colnames(support_matrix) != "(Intercept)"]
R> 
R> # Create test and train dataset
R> support_test <- support_matrix[indiv,,drop = FALSE]
R> 
R> support_train <- support_matrix[-indiv,]
R> y <- as.matrix(support[-indiv, c("death", "d.time")])
\end{CodeInput}
\end{CodeChunk}

We are now ready to estimate the hazard function. We use lasso
penalization; this is achieved by setting \code{alpha} to 1. Also, note
that the argument \code{formula_time} can be used to specify the
function form for time; in our example, we model time linearly.

@@

\begin{CodeInput}
R> support_cb <- fitSmoothHazard.fit(support_train, y, 
R+                                   formula_time = ~ d.time, 
R+                                   family = "glmnet", 
R+                                   event = "death", time = "d.time", 
R+                                   alpha = 1, ratio = 100)
\end{CodeInput}
\end{CodeChunk}

Then, we take our model object and use the \code{absoluteRisk} function
to compute our estimate of the cumulative incidence function. Since
\code{glmnet} fitted the penalized logistic regression for a range of
penalty terms, we need to specify which one we want to use for our
prediction. We select \code{s = "lambda.1se"}, as recommended.

@@

\begin{CodeInput}
R> time_points <- seq(1, max(support$d.time), by = 0.5)
R> support_risk <- absoluteRisk(support_cb,
R+                              time = time_points,
R+                              newdata = support_test,
R+                              s = "lambda.1se", method = "numerical")
\end{CodeInput}
\end{CodeChunk}

We will compare this curve to a regularized version of Cox regression,
and a Kaplan-Meier survival curve. Using the \pkg{glmnet} package, we
can fit the Cox regression model using regularized estimation.

@@

\begin{CodeInput}
R> # Create survival object for glmnet
R> surv_glmnet <- survival::Surv(time = y[, "d.time"], 
R+                               event = y[, "death"])
R> 
R> support_cox <- glmnet::cv.glmnet(x = support_train, 
R+                                  y = surv_glmnet,
R+                                  family = "cox", alpha = 1)
\end{CodeInput}
\end{CodeChunk}

However, \pkg{glmnet} does not provide any functionality for estimating
the baseline hazard, and the functionalities in the \pkg{survival}
package do not work on the output from \code{cv.glmnet}. To circumvent
this issue and get an estimate of the baseline hazard, we followed the
following steps. First, we fitted a non-regularized Cox model using only
the variables selected by \pkg{glmnet}; the baseline hazard will be
estimated using this object. Then, we replaced the estimates of the
regression coefficients by the estimates from \pkg{glmnet}. Finally,
this modified \code{coxph} object is fed to the \code{survival::survfit}
function to obtain the estimated survival function.

\begin{CodeInput}
R> # Identifying selected covariates
R> nonzero_covariates_cox <- predict(support_cox, type = "nonzero", 
R+                                   s = "lambda.1se")
R> nonzero_coef_cox <- coef(support_cox, s = "lambda.1se")
R> # Creating a new dataset that only contains the covariates selected by glmnet
R> support_cox_data <- as.data.frame(cbind(y,
R+                                         support_train[, nonzero_covariates_cox$X1]))
R> # Fitting a cox model
R> support_cox <- survival::coxph(Surv(time = d.time, event = death) ~ ., 
R+                           data = support_cox_data)
R> # Replace the estimate with those obtained by glmnet
R> support_cox$coefficients <- nonzero_coef_cox@x
R> 
R> # Fitting CI curve for cox+glmnet
R> support_cox_risk <- survival::survfit(support_cox,
R+                                       newdata = as.data.frame(support_test),
R+                                       time = time_points,
R+                                       type = "breslow")
\end{CodeInput}
\end{CodeChunk}

We used the \pkg{survival} package to calculate the Kaplan-Meier
estimate of the survival function.

\begin{CodeInput}
R> km <- survival::Surv(time = support$d.time, 
R+                      event = support$death)
R> abKm <- survival::survfit(km ~ 1,
R+                           type = 'kaplan-meier')
\end{CodeInput}
\end{CodeChunk}

Now that our three cumulative incidence functions have been calculated,
we compare them all in Figure \ref{fig:abSC}. Both \pkg{glmnet} and
\pkg{casebase} have a lower estimated cumulative incidence by the end of
the study, when compared to Kaplan-Meier.
\textcolor{red}{It would be good to have an idea of why the case-base and Coxnet curves are so different. At the very least, I think we should compare the select covariates.}

\begin{CodeChunk}
\begin{figure}

@@

\caption{\label{fig:abSC} Comparison of cumulative incidence curves between regularized casebase in black, regularized Cox regression in red and Kaplan-Meier in blue.}\label{fig:abSupportComparison}
\end{figure}
\end{CodeChunk}

@@

Note that the fourth model (i.e.~\texttt{fit4}) includes an interaction
term between exposure and follow-up time. In other words, this model no
longer exhibit proportional hazards. The evidence of non-proportionality
of hazards in the Stanford Heart Transplant data has been widely
discussed \citep{arjas1988graphical}.

@@

This report was generated on 2019-12-12 15:16:14 using the following
computational environment and dependencies:

@@

#> ─ Session info ──────────────────────────────────────────────────────────

@@

#>  tz       America/Winnipeg            
#>  date     2019-12-12                  

@@

#> ─ Packages ──────────────────────────────────────────────────────────────
#>  package     * version    date       lib

@@

#>  callr         3.3.1      2019-07-18 [1]
#>  casebase    * 0.2.1.9001 2019-12-11 [1]

@@

#>  cli           2.0.0      2019-12-09 [1]
#>  codetools     0.2-16     2018-12-24 [4]

@@

@@

#>  devtools      2.1.0      2019-07-06 [1]

@@

@@

#>  fansi         0.4.0      2018-10-05 [1]

@@

#>  foreach     * 1.4.4      2017-12-12 [1]

@@

#>  glmnet      * 2.0-18     2019-05-20 [1]

@@

#>  haven         2.1.1      2019-07-04 [1]

@@

#>  htmltools     0.3.6      2017-04-28 [1]

@@

#>  iterators     1.0.9      2017-12-12 [1]

@@

#>  knitr         1.23       2019-05-18 [1]

@@

#>  lattice       0.20-38    2018-11-04 [4]

@@

#>  Matrix      * 1.2-18     2019-11-27 [4]

@@

#>  mgcv          1.8-24     2018-06-18 [1]

@@

#>  nlme          3.1-142    2019-11-07 [4]
#>  pillar        1.4.2      2019-06-29 [1]
#>  pkgbuild      1.0.3      2019-03-20 [1]

@@

@@

#>  rmarkdown     1.16       2019-10-01 [1]

@@

#>  rticles       0.9.1      2019-07-22 [1]

@@

#>  stringi       1.4.3      2019-03-12 [1]

@@

#>  survival    * 2.44-1.1   2019-04-01 [4]
#>  testthat      2.2.0      2019-07-22 [1]

@@

#>  tidyverse   * 1.2.1      2017-11-14 [1]

@@

#>  vctrs         0.2.0      2019-07-05 [1]

@@

#>  webshot       0.5.1      2018-09-28 [1]

@@

#>  xfun          0.8        2019-06-25 [1]

@@

#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          

@@

#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.2)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.2)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.5.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.5.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.2)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.6.1)                          
#>  Github (rstudio/rticles@8d56fc6)        
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.2)                          
#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.1)                          

@@

#> [1] /home/mturgeon/Rlibs
#> [2] /usr/local/lib/R/site-library
#> [3] /usr/lib/R/site-library
#> [4] /usr/lib/R/library

@@

\begin{CodeOutput}
#> Local:    review /home/mturgeon/Documents/git_repositories/cbpaper
#> Head:     [c095709] 2019-12-12: review 2nd case study
\end{CodeOutput}
\end{CodeChunk}

>>>>

At the time of writing, a cursory examination of CRAN's task view on
survival analysis reveals that there are over 250 packages related to
survival analysis \citeyearpar{survTaskView}. For the purposes of this
article, we restricted our description to packages that implement at
least one of the following features: parametric modeling,
non-proportional hazard models, competing risk analysis, penalized
estimation, and cumulative incidence curve estimation. By searching for
appropriate keywords in the \code{DESCRIPTION} file of these packages,
we found 60 relevant packages. These 60 packages were then manually
examined to determine which ones are comparable to \pkg{casebase}. In
particular, we excluded packages that were focused on a different set of
problems, such as frailty and multi-state models. The remaining 14
packages appear in Table \ref{tab:surv-pkgs}, along with some of the
functionalities they offer.

Several packages implement penalized estimation for the Cox model:
\pkg{glmnet} \citeyearpar{regpathcox}, \pkg{glmpath}
\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
\pkg{RiskRegression} \citeyearpar{gerds_blanche}. Moreover, some
packages also include penalized estimation in the context of Cox models
with time-varying coefficients: elastic-net penalization with
\pkg{CoxRidge} \citeyearpar{perperoglou} and \pkg{rstpm2}
\citeyearpar{clements_liu}, while \pkg{survival}
\citeyearpar{survival-package} has an implementation for ridge.
\pkg{casebase} provides penalized estimation in the context of
parametric hazards. To our knowledge, \pkg{casebase} and \pkg{rsptm2}
are the only packages to offer this functionality.
\textcolor{red}{I can dive deeper to determine if this is a fair statement}

Parametric survival models are implemented in a handful of packages:
\pkg{CFC} \citeyearpar{mahani2015bayesian}, \pkg{flexsurv}
\citeyearpar{flexsurv}, \pkg{SmoothHazard} \citeyearpar{smoothHazard},
\pkg{rsptm2} \citeyearpar{clements_liu}, \pkg{mets}
\citeyearpar{scheike2014estimating}, and \pkg{survival}. The types of
models they allow vary for each package. For example, \pkg{SmoothHazard}
is limited to Weibull distributions \citeyearpar{smoothHazard}, whereas
both \pkg{flexsurv} and \pkg{survival} allow users to supply any
distribution of their choice. Also, \pkg{flexsurv}, \pkg{smoothhazard},
\pkg{mets} and \pkg{rstpm2} also have the ability to model the effect of
time using splines, which allows flexible modeling of the hazard
function. Moreover, \pkg{flexsurv} has the ability to estimate both
scale and shape parameters for a variety of parametric families (see
Table \ref{tab:dists}). As discussed above, \pkg{casebase} can model any
parametric family whose log-hazard can be expressed as a linear model of
covariates (including time). Therefore, our package allows the user to
model the effect of time using splines, and through interaction terms
involving covariates and time, it also allows user to fit time-varying
coefficient models. However, we do not explicitly model any shape
parameter, unlike \pkg{flexsurv}.

Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv},
\pkg{mets}, \pkg{cmprsk} and \pkg{survival} contain implementations for
competing risks. The differentiating factor between these packages and
\pkg{casebase} is that \pkg{casebase} handles competing risks through
multiple logistic regression.
\textcolor{red}{The discussion for \pkg{CFC} discusses aspects that I'm not certain are important for our argument. Mixture models, Vertical modelling,Psuedo-observations, machine learning models. Should i include these?}

@@

\begin{CodeInput}
R> 
R> data(bmtcrr)
\end{CodeInput}
\end{CodeChunk}

@@

For the third case study, we will look at regularized estimation for the
hazard function. This can be performed using a combination of case-base
sampling and regularized logistic regression. To demonstrate this
capability, we examine the Study to Understand Prognoses Preferences
Outcomes and Risks of Treatment (SUPPORT) dataset
\citep{knaus1995support}. The SUPPORT dataset tracks death for
individuals who are considered seriously ill within a hospital.
Imputation was conducted using the \pkg{mice} package in \proglang{R}
with default settings \citep{mice}. Before imputation, there were a
total of 1000 individuals and 35 variables. After imputation, we have
690 individuals.
\textcolor{red}{we lose individuals due to coliniarity. If a pair of covariates are deemed colinear to a certain degree, they are excluded from any imputations. here, only edu  income  avtisst*  race  ph*  glucose*  adlp* were imputed.}
Of these, 66.09\% experienced the event of interest. A description of
each variable can be found in Table \ref{tab:support1} and a breakdown
of each categorical variable in Table \ref{tab:support2}. The original
data was obtained from the Department of Biostatistics at Vanderbilt
University
\textcolor{red}{do we need to reference the dataset itself? Or is the paper enough? In my code for generating the data in casebase, i link to the original data online and pull from there}.
The imputed dataset is available as part of the \pkg{casebase} package.

@@

As before, the first step is to visualize the incidence density. The
population-time plot for the SUPPORT data appears in Figure
\ref{fig:support-poptime}. \textbackslash{}textcolor\{red\}\{red dots
now go all over. percentile\_number introduced as a parameter in
popTime\}

@@

\begin{CodeInput}
R> data(support)
R> support_popTime <- popTime(support,
R+                            time = "d.time", event = "death",percentile_number = 0.31)
R> plot(support_popTime)
\end{CodeInput}
\begin{figure}

@@

Our main objective is to compute the absolute risk of death for a given
set of covariates, using regularization when fitting our model. First,
we fit a smooth hazard to the data; for the sake of this example, we
opted for a linear term for time. As casebase makes use of the glmnet
package, we interact with the fitsmoothhazard.fit function using a
matrix interface, where y contains the time and event variables and x
contains all other variables we would like to include in the model. For
the SUPPORT dataset, all factor variables were converted into dummy
variables. Here, we used lasso penalization by setting alpha to 1.

@@

\begin{CodeInput}
R> #Using casebase's glmnet implementation, using splines and a case to base ratio of 100,
R> #to determine the hazard function
R> cb.Model=casebase::fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
R+                                        formula_time = ~d.time,alpha=1,ratio=100)
\end{CodeInput}
\end{CodeChunk}

Then, we take our model and use the absoluteRisk function to integrate
and retrieve our absolute risk function.

@@

\begin{CodeInput}
R> #Estimating the absolute risk curve using the newData parameter.
R> casebaseAbsolute=casebase::absoluteRisk(cb.Model,time = seq(1,max(completeData$d.time), 50),newdata = newData , s="lambda.1se",method=c("numerical"))
R> cab=data.frame(casebaseAbsolute[,1],rowMeans(casebaseAbsolute[,-c(1)]))
\end{CodeInput}
\end{CodeChunk}

We will compare this curve to a regularized version of cox regression,
and a kaplan-meier survival curve. The cox regression absolute risk
curve with lasso penalization required two extra steps to retrieve.
Coxnet from the glmnet package has tools to fit a regularized cox model,
but requires that a survival object skeleton of the selected variables,
so that the survival package tools can be used to retrive the absolute
risk. We handle this by fitting a regularized cox model with the glmnet
package, and a cox model from the survival package with the selected
variables from the regularized fit. This second model serves as a
skeleton that permits the use of survival package functions. The
coefficients from the regularized model will replace the coeficients in
our skeleton. the resulting model will have the correct coefficients
when integrating, but will have an incorrect standard error. For the
purposes of this case study, we are only interested in the absolute risk
curve itself and not the standard error.

@@

\begin{CodeInput}
R> #Create u and xCox, which will be used when fitting Cox with glmnet.
R> u=survival::Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))
R> xCox=as.matrix(sparse.model.matrix(death~ .-d.time,data=completeData))
R> #hazard for cox using glmnet
R> coxglmFit=glmnet::cv.glmnet(x=xCox,y=u, family="cox",alpha=1)
R> #convergence demonstrated in plot
R> plot(coxglmFit)
\end{CodeInput}

\begin{center}\includegraphics{../figures/coxHazAbsolute-1} \end{center}

\begin{CodeInput}
R> #taking the coefficient estimates for later use
R> nonzero_covariate_cox <- predict(coxglmFit, type = "nonzero", s = "lambda.1se")
R> nonzero_coef_cox <- coef(coxglmFit, s = "lambda.1se")
R> #creating a new dataset that only contains the covariates chosen through glmnet.
R> #cleanCoxData<- as.data.frame(cbind(as.numeric(workingData$d.time),as.factor(workingData$death), xCox[,nonzero_covariate_cox$X1]))
R> cleanCoxData<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCox[,nonzero_covariate_cox$X1]))
R> #newDataCox<-xCox[sam,nonzero_covariate_cox$X1]
R> #fitting a cox model using regular estimation, however we will not keep it.
R> #this is used more as an object place holder.
R> coxFit <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxData)
R> #The coefficients of this object will be replaced with the estimates from coxglmFit.
R> #Doing so makes it so that everything is invalid aside from the coefficients.
R> #In this case, all we need to estimate the absolute risk is the coefficients.
R> #Std. error would be incorrect here, if we were to draw error bars.
R> coxFit$coefficients<-nonzero_coef_cox@x
R> #Fitting absolute risk curve for cox+glmnet. -1 to remove intercept
R> newDataCox=newData[,nonzero_covariate_cox$X1-1]
R> 
R> abCoxFit<-survival::survfit(coxFit,newdata=as.data.frame(newDataCox),time = seq(0,max(completeData$d.time), 1),type="breslow")
R> 
R> abCF<-abCoxFit
R> abCF$surv<-rowMeans(abCF$surv)
\end{CodeInput}
\end{CodeChunk}

We used the survival package to calculate the kaplan-meier absolute risk
function.

@@

\begin{CodeInput}
R> #creating a surv object to be used to fit an unadjusted absolute risk curve.
R> # (Kaplan-Meier risk curve)
R> km <- survival::Surv(time = completeData$d.time, event = completeData$death)
R> abKm<-survival::survfit(km~1,type='kaplan-meier',conf.type='log')
\end{CodeInput}
\end{CodeChunk}

Now that our three absolute risk functions have been calculated, we
compare them all in Figure \ref{fig:abSC}. Both coxnet and casebase
decrease the absolute risk by the end of the study, in comparison to
kaplan-meier.

@@

\begin{CodeInput}
R> # A plot to compare all three Absolute risk (Commulative Incidence)
R> plot(cab[,1],cab[,2],type='l',col="black",lwd=3,main="Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves",xlab="Survival-Time",ylab="Cumulative Incidence",ylim=c(0,1),xlim=)
R> lines(abCF,col="red",fun="event",lwd=3,conf.int = FALSE)
R> lines(abKm ,col="Blue",fun="event",lwd=3,conf.int = FALSE)
R> legend("bottomright", 
R+        legend = c( "Casebase (Lasso+linear)","semi-parametric (Cox)+glmnet","KM curve"), 
R+        col = c("black","red","Blue"),
R+        lty = c(1, 1, 1), 
R+        bg = "gray90")
\end{CodeInput}
\begin{figure}

@@

\caption{\label{fig:abSC} Compares regularized casebase in black, regularized cox in red and Kaplan meier in blue.}\label{fig:abSupportComparison}
\end{figure}
\end{CodeChunk}

@@

Note that the fourth model (i.e. \texttt{fit4}) includes an interaction
term between exposure and follow-up time. In other words, this model no
longer exhibit proportional hazards. The evidence of non-proportionality
of hazards in the Stanford Heart Transplant data has been widely
discussed \citep{arjas1988graphical}.

@@

This report was generated on 2020-01-16 13:51:03 using the following
computational environment and dependencies:

@@

#> - Session info ---------------------------------------------------------------

@@

#>  tz       America/Montreal            
#>  date     2020-01-16                  

@@

#> - Packages -------------------------------------------------------------------
#>  ! package     * version    date       lib

@@

#>    callr         3.4.0      2019-12-09 [1]
#>  P casebase    * 0.2.1.9001 2019-12-11 [?]

@@

#>    cli           2.0.1      2020-01-08 [1]
#>    codetools     0.2-16     2018-12-24 [1]

@@

#>    DBI           1.0.0      2018-05-02 [1]
#>    dbplyr        1.4.2      2019-06-17 [1]

@@

#>    devtools      2.2.1      2019-09-24 [1]

@@

#>    ellipsis      0.3.0      2019-09-20 [1]

@@

#>    fansi         0.4.1      2020-01-08 [1]

@@

#>    foreach       1.4.7      2019-07-27 [1]

@@

#>    glmnet      * 3.0-1      2019-11-15 [1]

@@

#>    haven         2.2.0      2019-11-08 [1]

@@

#>    htmltools     0.4.0      2019-10-04 [1]

@@

#>    iterators     1.0.12     2019-07-26 [1]

@@

#>    knitr         1.26       2019-11-12 [1]

@@

#>    lattice       0.20-38    2018-11-04 [1]

@@

#>    Matrix      * 1.2-17     2019-03-22 [1]

@@

#>    mgcv          1.8-31     2019-11-09 [1]

@@

#>    nlme          3.1-142    2019-11-07 [1]
#>    pillar        1.4.3      2019-12-20 [1]
#>    pkgbuild      1.0.6      2019-10-09 [1]

@@

#>    reprex        0.3.0      2019-05-16 [1]

@@

#>    rmarkdown     1.18       2019-11-27 [1]

@@

#>    rticles       0.12       2019-11-12 [1]

@@

#>    shape         1.4.4      2018-02-07 [1]
#>    stringi       1.4.5      2020-01-11 [1]

@@

#>    survival    * 3.1-7      2019-11-09 [1]
#>    testthat    * 2.3.1      2019-12-01 [1]

@@

#>    tidyverse   * 1.3.0      2019-11-21 [1]

@@

#>    vctrs         0.2.1      2019-12-17 [1]

@@

#>    webshot       0.5.2      2019-11-22 [1]

@@

#>    xfun          0.11       2019-11-12 [1]

@@

#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          

@@

#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          

@@

#> [1] /Library/Frameworks/R.framework/Versions/3.6/Resources/library
#>  P -- Loaded and on-disk path mismatch.

@@

\begin{CodeOutput}
#> Local:    master /Users/jislam/Desktop/Gits/cbpaper
#> Remote:   master @ origin (https://github.com/Jesse-Islam/cbpaper.git)
#> Head:     [a528002] 2019-12-12: Merge pull request #12 from sahirbhatnagar/review
\end{CodeOutput}
\end{CodeChunk}
